Shapes analysis for time series.

### 2024-05-22 NEURIPS24/appendix.tex
final

\section{Appendix for experiment: TS-LDDMM representation identifiability}
\subsection{Settings}
\label{appendix: settings_identifiability}
This experiment only involves the TS-LDDMM method in two different settings: 
\begin{itemize}
  \item \textbf{The velocity field kernel $K_G$ is well-specified:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,100,1,1)$, the varifold loss kernels $(k_{pos},k_{dir})$ are set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (2,1,2,0.6)$, and the optimizer has 400 steps with a maximum stepsize $\eta_M$ of 0.05.
  \item \textbf{The velocity field kernel $K_G$ is missspecified:} The velocity field kernel $K_G$ is set with  $(c_0,c_1,\sigma_{T,1}) = (1,0.1,1)$, $\sigma_{T,0}$ ranging in $(1,5,10,50,100,200,300)$, and $\sigma_x$  ranging in $(0.1,1,10,100)$. The varifold loss kernels $(k_{pos},k_{dir})$ are set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (2,1,2,0.6)$, and the optimizer has 400 steps with a maximum stepsize $\eta_M$ of 0.05.
\end{itemize}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\linewidth]{pictures/samples.pdf}
  \vspace{-1em}
  \caption{Plots of $\varphi^{\{v_0(\mathbf{\alpha}^*,\msx)\}}.\msx$ for different values of $\mathbf{\alpha}^*$ according to its sampling parameter $t_a,s_a,m_s $, taking $\msx=\msg(s_0)$ with $s_0:k\in [300]\to \sin(2\pi k/300) $.}
  \label{fig:exemple_synthetic}
  \vspace{-1em}
\end{figure}

\begin{table}
  \caption{Values of $\scrl(\varphi^{\{v_0(\mathbf{\alpha}^*,\msx)\}}.\msx,\varphi^{\{\hat{v}_0\}}.\msx)$ as $\mathbf{\alpha}^*$ is sampled according to Gen(10,10,50) and $\hat{v}_0$ is estimated using $K_\msg$ with varying parameters $\sigma_{T,1},\sigma_x$.}
    \centering
       \begin{tabular}{lrrrrrrr}
       \toprule
       $\sigma_{T,0} \backslash \sigma_x$  & 1 & 10 & 50 & 100 & 200 & 300 \\
       \midrule
       0.1 & 2e+0 & 3e-4  & 1e-5&4e-6&7e-4&4e-3 \\
      1 & 4e-2 & 1e-4  & 1e-5&4e-6&7e-4 &4e-3  \\
       100 & 4e-2 & 2e-4  & 1e-5&4e-6&7e-4&4e-3  \\
       \bottomrule
       \end{tabular}
    \label{table:synthetic2}
    \vspace{-1em}
\end{table}

@@

First, we show the model identifiability when the kernel $K_G$ is well specified: the estimated parameter is a good approximation of the generating parameter when the generation and the estimation procedure use the same hyperparameters for the RKHS kernel $K_\msg$.
`%All the hyperparameter values for generation and estimation are given in \Cref{appendix:numerics_synthetic}.
We fix the initial control points as $\msx=\left(x_k=(k,\sin(2\pi k/300))\right)_{k\in[300]} $.
Given $m_{s}\in \Nset_{>0}$ and $t_{a},s_{a}>0$, we randomly generate initial momentums $\mathbf{\alpha}^*=(\alpha_k^*)_{k\in[\mathbf{n}_0]}$ with the following sampling, called Gen($m_s,t_a,s_a$):
For any $k\in[\mathbf{n}_0]$, $\mathbf{\alpha}_k'$ is sampled according to a Gaussian normal distribution $\mathcal{N}(0_{d+1},I_{d+1})$.
Then, $(\alpha_k')_{k\in[\mathbf{n}_0]}$ is regularized by a rolling average of size $m_{s}$, we get $\bar{\mathbf{\alpha}}'=(\bar{\alpha}_k')_{k\in[\mathbf{n}_0]}$.
Finally, we normalize $\bar{\mathbf{\alpha}}'$ to derive $\mathbf{\alpha}^*$ such that $|([\alpha_k^*]_t)_{k\in[\mathbf{n}_0]}|=t_{\text{amp}}$ and $|([\alpha_k^*]_s)_{k\in[\mathbf{n}_0]}|=s_{\text{amp}}$ for any $k\in[\mathbf{n}_0]$, denoting by $[\alpha_k^*]_t,[\alpha_k^*]_s$ the time and space coordinates of $\alpha_k^*$ respectively.
Note that the regularizing step $(\mathbf{\alpha}_k')_{k\in[\mathbf{n}_0]}\to \bar{\mathbf{\alpha}}' $ is necessary to obtain realistic deformations which take into account the regularity induced by the RKHS $\msv$.

@@

Secondly, we demonstrate a weak identifiability when the kernel $K_\msg$ is misspecified: we can reconstruct the graph time series' after deformations even if the hyperparameters of $K_\msg$
are different during the generation and the estimation.
 The hyperparameters of $K_\msg$ during generation are $(c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x)=(1,0.1,100,1,1)$ and we fix $\sigma_{T,1},c_0,c_1=(1,1,0.1) $ for $K_\msg$ during estimation.
 We aim to understand the impact of $\sigma_{T,1},\sigma_x$ on the reconstruction since they are encoding the smoothness of the transformation according to time and space.

@@

\subsection{Protocol}
In this experiment, we investigate the robustness to missing samples and the classification performance of TS-LDDMM compared to Neural ODEs on 15 datasets described in \Cref{appendix:classification_dataset}. For fairness between methods of different architectures, the evaluation protocol on each dataset and method is as follows: 
\begin{enumerate}
  \item The dataset is spilt in train 75\%, validation 15\%, and test 15\%.
  \item hyperparameters are tuned with train and validation sets and a missing rate of 0\%.
  \item For each missing rate in [0\%,30\%,50\%,70\%]
    \begin{itemize}
      \item Remove samples in time series in the train and test sets according to the missing rate and the drop procedure described in~\cite{kidger2020neural}.
      \item Train the model on the train set
      \item Evaluate the macro f1-score on the test set
    \end{itemize}
\end{enumerate}

@@

\subsection{Benchmark metohds}
\begin{itemize}
  \item SRV-based method: we include TCLR~\cite{heo2024logistic} a logistic regression on the tangent space of the Frechet mean with Square Root Velocity (SRV representation). We also include Shape-FPCA~\cite{wu2024shape} that encodes both the time series and its time parameterization. Both methods are implemented with the Python library \texttt{FDASRSF}~\footnote{https://fdasrsf-python.readthedocs.io/en/latest/}, which cannot handle multivariate time series.
  \item LDDMM-Based : TS-LDDMM (ours) and LDDMM~\cite{glaunes2008large}. Both methods learn representations by solving ODEs parametrized with Kernels. While both methods handle multivariate signals, TS-LDDMM is specifically designed for time series.
 \end{itemize}
\subsection{Settings}
\label{appendix: settings_classification}

%\paragraph{Methods} We compare TS-LDDMM with a method from function~\cite{wu2024shape}

%\paragraph{Protocole}

\begin{table}[hbt!]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{llrrrr}
    \toprule
     & \textbf{Dataset} & \textbf{Shape-FPCA (2024)} & \textbf{TCLR (2024)} & \textbf{LDDMM (2008)} & \textbf{TS-LDDMM (ours)} \\
    \midrule
    \multirow[c]{7}{*}{Univariate} & ArrowHead & 0.18 & 0.75 & \underline{0.84} & \textbf{0.91} \\
     & BME & 0.16 & \underline{1.00} & 0.82 & \textbf{1.00} \\
     & ECG200 & 0.40 & 0.67 & \textbf{0.81} & \underline{0.79} \\
     & FacesUCR & 0.08 & \underline{0.73} & 0.69 & \textbf{0.86} \\
     & GunPoint & 0.93 & \underline{0.97} & 0.83 & \textbf{1.00} \\
     & PhalangesOutlinesCorrect & 0.39 & \textbf{0.63} & \underline{0.53} & 0.52 \\
     & Trace & 0.55 & \underline{1.00} & 0.46 & \textbf{1.00} \\
    \cline{1-6}
    \multirow[c]{8}{*}{Multivariate} & ArticularyWordRecognition & -- & -- & \underline{0.98} & \textbf{1.00} \\
     & Cricket & -- & -- & \underline{0.77} & \textbf{0.93} \\
     & ERing & -- & -- & \underline{0.95} & \textbf{0.98} \\
     & Handwriting & -- & -- & \underline{0.22} & \textbf{0.44} \\
     & Libras & -- & -- & \underline{0.56} & \textbf{0.60} \\
     & NATOPS & -- & -- & \underline{0.82} & \textbf{0.82} \\
     & RacketSports & -- & -- & \textbf{0.83} & \underline{0.79} \\
     & UWaveGestureLibrary & -- & -- & \underline{0.72} & \textbf{0.81} \\
    \bottomrule
    \end{tabular}
    %
  }

>>>>

\section{Appendix for experiment: TS-LDDMM representation identifiability}
In this experiment, we evaluate the ability of TS-LDDMM to retrieve the parameter $v_0^*$ that encodes the deformation $\varphi^{\{v_0^*\}}$ acting on a time series graph $\msg$ by solving the geodesic shooting problem \eqref{eq:relaxation} between $\msg$ and $\varphi^{\{v_0^*\}}.\msg$. Parameter identifiability is an important property for subsequent statistical analysis. Results show that TS-LDDMM representations are identifiable or weakly identifiable depending on the velocity field kernel $K_G$ specification.

\subsection{Settings}
\label{appendix: settings_identifiability}
This experiment only involves the TS-LDDMM method in two different settings: 
\begin{itemize}
  \item \textbf{The velocity field kernel $K_G$ is well-specified:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,100,1,1)$, the varifold loss kernels $(k_{pos},k_{dir})$ are set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (2,1,2,0.6)$, and the optimizer has 400 steps with a maximum stepsize $\eta_M$ of 0.05.
  \item \textbf{The velocity field kernel $K_G$ is missspecified:} The velocity field kernel $K_G$ is set with  $(c_0,c_1,\sigma_{T,1}) = (1,0.1,1)$, $\sigma_{T,0}$ ranging in $(1,5,10,50,100,200,300)$, and $\sigma_x$  ranging in $(0.1,1,10,100)$. The varifold loss kernels $(k_{pos},k_{dir})$ are set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (2,1,2,0.6)$, and the optimizer has 400 steps with a maximum stepsize $\eta_M$ of 0.05.
\end{itemize}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\linewidth]{pictures/samples.pdf}
  \caption{Plots of $\varphi^{\{v_0(\mathbf{\alpha}^*,\msx)\}}.\msx$ for different values of $\mathbf{\alpha}^*$ according to its sampling parameter $t_a,s_a,m_s $, taking $\msx=\msg(s_0)$ with $s_0:k\in [300]\to \sin(2\pi k/300) $.}
  \label{fig:exemple_synthetic}
\end{figure}

\begin{table}
  \caption{Values of $\scrl(\varphi^{\{v_0(\mathbf{\alpha}^*,\msx)\}}.\msx,\varphi^{\{\hat{v}_0\}}.\msx)$ as $\mathbf{\alpha}^*$ is sampled according to Gen(10,10,50) and $\hat{v}_0$ is estimated using $K_\msg$ with varying parameters $\sigma_{T,1},\sigma_x$.}
    \centering
       \begin{tabular}{lrrrrrrr}
       \toprule
       $\sigma_{T,0} \backslash \sigma_x$  & 1 & 10 & 50 & 100 & 200 & 300 \\
       \midrule
       0.1 & 2e+0 & 3e-4  & 1e-5&4e-6&7e-4&4e-3 \\
      1 & 4e-2 & 1e-4  & 1e-5&4e-6&7e-4 &4e-3  \\
       100 & 4e-2 & 2e-4  & 1e-5&4e-6&7e-4&4e-3  \\
       \bottomrule
       \end{tabular}
    \label{table:synthetic2}
\end{table}

@@

\paragraph{The velocity field kernel $K_G$ is well specified.}

First, we show the model identifiability when the kernel $K_G$ is well specified: the estimated parameter is a good approximation of the generating parameter when the generation and the estimation procedure use the same hyperparameters for the RKHS kernel $K_\msg$.
All the hyperparameter values for generation and estimation are given in \Cref{appendix: settings_identifiability}.

We fix the initial control points as $\msx=\left(x_k=(k,\sin(2\pi k/300))\right)_{k\in[300]} $.
Given $m_{s}\in \Nset_{>0}$ and $t_{a},s_{a}>0$, we randomly generate initial momentums $\mathbf{\alpha}^*=(\alpha_k^*)_{k\in[\mathbf{n}_0]}$ with the following sampling, called Gen($m_s,t_a,s_a$):
For any $k\in[\mathbf{n}_0]$, $\mathbf{\alpha}_k'$ is sampled according to a Gaussian normal distribution $\mathcal{N}(0_{d+1},I_{d+1})$.
Then, $(\alpha_k')_{k\in[\mathbf{n}_0]}$ is regularized by a rolling average of size $m_{s}$, we get $\bar{\mathbf{\alpha}}'=(\bar{\alpha}_k')_{k\in[\mathbf{n}_0]}$.
Finally, we normalize $\bar{\mathbf{\alpha}}'$ to derive $\mathbf{\alpha}^*$ such that $|([\alpha_k^*]_t)_{k\in[\mathbf{n}_0]}|=t_{\text{amp}}$ and $|([\alpha_k^*]_s)_{k\in[\mathbf{n}_0]}|=s_{\text{amp}}$ for any $k\in[\mathbf{n}_0]$, denoting by $[\alpha_k^*]_t,[\alpha_k^*]_s$ the time and space coordinates of $\alpha_k^*$ respectively.
Note that the regularizing step $(\mathbf{\alpha}_k')_{k\in[\mathbf{n}_0]}\to \bar{\mathbf{\alpha}}' $ is necessary to obtain realistic deformations which take into account the regularity induced by the RKHS $\msv$.

@@

\paragraph{The velocity field kernel $K_G$ is misspecified.}

We demonstrate a weak identifiability when the kernel $K_\msg$ is misspecified: we can reconstruct the graph time series' after deformations even if the hyperparameters of $K_\msg$
are different during the generation and the estimation.
 The hyperparameters of $K_\msg$ during generation are $(c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x)=(1,0.1,100,1,1)$ and we fix $\sigma_{T,1},c_0,c_1=(1,1,0.1) $ for $K_\msg$ during estimation.
 We aim to understand the impact of $\sigma_{T,1},\sigma_x$ on the reconstruction since they are encoding the smoothness of the transformation according to time and space.

@@

\subsection{Protocol}

In this experiment, we investigate the robustness to missing samples and the classification performance of TS-LDDMM compared to Neural ODEs on 15 datasets described in \Cref{appendix:classification_dataset}. For fairness between methods of different architectures, the evaluation protocol on each dataset and method is as follows: 
\begin{enumerate}
  \item Spilt the dataset in train 75\%, validation 15\%, and test 15\%.
  \item Tune hyperparameters with train and validation sets and a missing rate of 0\%.
  \item For each missing rate in [0\%,30\%,50\%,70\%]
    \begin{itemize}
      \item Remove samples in time series in the train and test sets according to the missing rate and the drop procedure described in~\cite{kidger2020neural}.
      \item Train the model on the train set
      \item Evaluate the macro f1-score on the test set
    \end{itemize}
\end{enumerate}

@@

\subsection{Benchmark metohds}
\begin{itemize}
  \item SRV-based method: we include TCLR~\cite{heo2024logistic} a logistic regression on the tangent space of the Frechet mean with Square Root Velocity (SRV representation). We also include Shape-FPCA~\cite{wu2024shape} that encodes both the time series and its time parameterization. 
  \item LDDMM-Based : TS-LDDMM (ours) and LDDMM~\cite{glaunes2008large}. Both methods learn representations by solving ODEs parametrized with Kernels. While both methods handle multivariate signals, TS-LDDMM is specifically designed for time series.
 \end{itemize}

\subsection{Model settings}

\paragraph{TCLR \& Shape-FPCA}
Shape-FPCA is available in the Python library \texttt{FDASRSF}~\footnote{https://fdasrsf-python.readthedocs.io/en/latest/}. Once the shape-FPCA representations are learned, they are fed to an \href{https://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html#sklearn.svm.SVC}{SVC} from \texttt{scikit-learn}. \texttt{FDASRSF} provides SRV representation methods that we combined with a \href{https://scikit-learn.org/stable/modules/generated/sklearn.linear_model.LogisticRegressionCV.html}{logistic regression} from \texttt{scikit-learn} to implement TCLR. For both methods, the number of steps to learn the Frechet mean is set to 50, and the regularization hyperparameter C is set through grid search on a validation set with the macro f1-score. Other parameters are set to default.

\paragraph{TS-LDDMM \& LDDMM}
Representations learned with TS-LDDMM or LDDMM are fed to an SVC from \texttt{scikit-learn}. All SVC's hyperparameters are set to default except the regularization term C, which is set through grid search on a validation set with the macro f1-score.

\paragraph{Protocole}
For each dataset and method, the evaluation protocol is a simple train,validation test with hyperparameter tuning:
\begin{enumerate}
  \item Split The dataset in train 75\%, validation 15\%, and test 15\%.
  \item Training and hyperparameters tuning with train and validation sets
  \item Evaluate the macro f1-score on the test set 
\end{enumerate}

\subsection{Results}
In this experiment, we investigate the classification performances of several methods from shape analysis on 15 shape-based time series datasets (7 univariate and 8 multivariate). The performances are evaluated in terms of macro f1-score. Results are aggregated in \Cref{table: shape classif}.

The TS-LDDMM-based classifier outperforms other methods on 12 out of 15 datasets. TCLR is the second-best performer on univariate datasets; however, its current implementation with \texttt{FDASRSF} does not extend to the multivariate case, which limits usage. LDDMM performances are lower than TCLR, and Shape-FPCA is the worst performer.

Overall, TS-LDDMM representations are well suited for shape-based time series classification, and its extension to multivariate irregularly sampled time series makes it a relevant option for time series shape analysis.

\begin{table}[hbt!]
  \centering
  \caption{F1-score comparison between methods from shape analysis on 15 datasets. \textbf{First} and \underline{second} best performers.}
  \label{table: shape classif}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{llrrrr}
    \toprule
     & \textbf{Dataset} & \textbf{Shape-FPCA (2024)} & \textbf{TCLR (2024)} & \textbf{LDDMM (2008)} & \textbf{TS-LDDMM (ours)} \\
    \midrule
    \multirow[c]{7}{*}{Univariate} & ArrowHead & 0.18 & 0.75 & \underline{0.84} & \textbf{0.91} \\
     & BME & 0.16 & \underline{1.00} & 0.82 & \textbf{1.00} \\
     & ECG200 & 0.40 & 0.67 & \textbf{0.81} & \underline{0.79} \\
     & FacesUCR & 0.08 & \underline{0.73} & 0.69 & \textbf{0.86} \\
     & GunPoint & 0.93 & \underline{0.97} & 0.83 & \textbf{1.00} \\
     & PhalangesOutlinesCorrect & 0.39 & \textbf{0.63} & \underline{0.53} & 0.52 \\
     & Trace & 0.55 & \underline{1.00} & 0.46 & \textbf{1.00} \\
    \cline{1-6}
    \multirow[c]{8}{*}{Multivariate} & ArticularyWordRecognition & -- & -- & \underline{0.98} & \textbf{1.00} \\
     & Cricket & -- & -- & \underline{0.77} & \textbf{0.93} \\
     & ERing & -- & -- & \underline{0.95} & \textbf{0.98} \\
     & Handwriting & -- & -- & \underline{0.22} & \textbf{0.44} \\
     & Libras & -- & -- & \underline{0.56} & \textbf{0.60} \\
     & NATOPS & -- & -- & \underline{0.82} & \textbf{0.82} \\
     & RacketSports & -- & -- & \textbf{0.83} & \underline{0.79} \\
     & UWaveGestureLibrary & -- & -- & \underline{0.72} & \textbf{0.81} \\
    \bottomrule
    \end{tabular}
    %
  }
